RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-10-29 formalframework.tex
added tool screenshots

Now that we have defined the semantics for RationalGRL, we can use this to define the notion of a valid RationalGRL model.

\begin{definition}[Valid RationalGRL Model]
\label{def:valid-rationalgrl-model}
A RationalGRL model $RatGRL = (Arg, Att)$ with preferred extension $E=(A_1,\ldots,A_n)$ is a \emph{valid RationalGRL model} iff the GRL model constructed from $E$ is a valid GRL model (Def.~\ref{def:valid-grl-model}).
\end{definition}

\todo{Marc}{Marc}{Perhaps say something about how to construct it. It is very trivial: Just remove the generic arguments and attack links, only keep the other arguments which directly translate to GRL elements and links.}

@@

In the previous subsection we formalized a \emph{static} representation of the RationalGRL framework. In this section we formalize the \emph{dynamics}. We do so by developing algorithms for applying argument schemes and critical questions in the context of a RationalGRL model (Def.~\ref{def:rationalgrl-model}). These algorithms are applied to RationalGRL models and produce new arguments and attack relations. We can then use argumentation semantics (Def.~\ref{def:semantics}) to compute sets of accepted arguments. The content of these arguments is then used to compute the resulting RationalGRL model, together with enabled and disabled GRL elements and their underlying arguments.

As discussion in Section~\ref{sect:overview}, ll the argument schemes and critical questions of Table~\ref{table:argument-schemes} fall into one of the following four categories:
\begin{itemize}
\item \emph{INTRO}: These algorithms add one or multiple new arguments to the RationalGRL model, without creating any attack links. The arguments are for new GRL elements or links.
\item \emph{DISABLE}: These algorithms add a single argument to the RationalGRL model, which attacks one of multiple existing arguments. The counterargument that is added does not correspond to any GRL element, but instead disables an existing one by attacking all arguments for it.
\item \emph{REPLACE}: These algorithms can be seen as a combination of \emph{INTRO} and \emph{DISABLE}. They add a new argument corresponding to a GRL element or link, and this new argument attacks a previous version of the element or link. 
\end{itemize}

@@

\paragraph{Rationale and example} Algorithm takes one argument, namely the name of the actor $a$. On line~\ref{alg:as0:1} of the algorithm, a new (unique) id is minted as the identifier of the new actor, which is assigned with its corresponding name to the variable $actor$. On line~\ref{alg:as0:2} the actor is added as a new argument.

Consider for instance...\todo{Marc}{Marc}{Add example}

\begin{algorithm}[h]
  \caption{AS1: Actor with id $i$ has resource $R$}\label{alg:add-actor}
  \begin{algorithmic}[1]
    \Procedure{$AS_1$}{$i, R$}
    \State $res\_id\gets mintId()$\label{alg:add-actor:id}
    \State $A_1\leftarrow (res\_id, Resource, R, AND)$\label{alg:add-actor:arg1}
    \State $A_2\leftarrow (mintId(),ActIE, i, res\_id)$\label{alg:add-actor:arg2}
    \State $Arg\gets Arg\cup \{A_1,A_2\}$\label{alg:add-actor:add-arg}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

@@

The algorithm itself runs as follows: on line~\ref{alg:add-actor:id}, a unique id is assigned to variable $res\_id$ (``resource identifier''). On line~\ref{alg:add-actor:arg1}, an argument for a resource with name $R$ and identifier $res\_id$ is assigned to $A_1$. On line~\ref{alg:add-actor:arg2}, a second argument $A_2$ is created, which is an argument for an Actor-IE relation (Def.~\ref{def:act-ie-relation}) between the input actor $i$ and the newly created element $res\_id$. Finally, on line~\ref{alg:add-actor:add-arg}, the arguments are added to the set of arguments $Arg$.

@@

\begin{algorithm}[h]
  \caption{AS5: Goal with id $i$ decomposes into task $T$}\label{alg:add-decomp}
  \begin{algorithmic}[1]
    \Procedure{$AS_5$}{$i, T$}
    \State $task\_id\gets mintId()$\label{alg:add-decomp:task-id}
    \State $A_1\leftarrow (task\_id, Task, t, AND)$\label{alg:add-decomp:arg1}
    \State $A_2\leftarrow (mintId(), Decomp, i, task\_id)$\label{alg:add-decomp:arg2}
    \State $Arg\gets Arg\cup \{A_1,A_2\}$\label{alg:add-decomp:add-arg}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

@@

@@

Consider for example...\todo{Marc}{Marc}{Add example here}

@@

\todo{Marc}{Marc}{Give example}

@@

\begin{algorithm}[h]
  \caption{CQ5c: Is the decomposition type of element $ie_i$ correct? No, it should be $X$ }\label{alg:replace1}
  \begin{algorithmic}[1]
    \Procedure{$CQ5c$}{$ie_i, X$}
    \State $A \leftarrow ie_i$\label{alg:replace1:arg}
    \State $A.decomptype\leftarrow X$\label{alg:replace1:decompchange}
    \State $IEArgs\leftarrow IE_i\subseteq  Arg$\label{alg:replace1:ieargs}
    \For{$A_i\in IEArgs$}\label{alg:replace1:for1}
      \State $Att \leftarrow Att \cup \{(A,A_i)\}$\label{alg:replace1:att}
    \EndFor
    \For{$\{(A_i,A_j)\in Att\mid A_j\in IEArgs\}$}\label{alg:replace1:for2}
      \State $Att\leftarrow (A_i,A)$
    \EndFor
    \For{$\{(A_i,A_j)\in Att\mid A_i\in IEArgs\}$}\label{alg:replace1:for3}
      \State $Att\leftarrow (A,A_j)$
    \EndFor
    \State $Arg\leftarrow Arg \cup \{A\}$\label{alg:replace1:addarg}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\paragraph{Rationale and example} While the original critical question CQ5c is specific to the decomposition between a goal and a task, Algorithm~\ref{alg:replace1} is more generally applicable to any IE, since all IEs have a decomposition type in their definition (Def.~\ref{def:ie}).

Let us go through this algorithm step by step. On line~\ref{alg:replace1:arg}, a new argument $A$ is created which is identical to original IE. Recall from Def.~\ref{def:notation} that we may refer to elements of an IE with the ``.'' notation. This is what we do on line~\ref{alg:replace1:decompchange}, where the decomposition type of the argument is changed to $X$. On line~\ref{alg:replace1:ieargs}, the set $IEArgs$ is assigned with all existing arguments for the input IE (recall the notation of $IE_i$ as the set of all IEs with id $i$, see Def.~\ref{def:notation}). Then, in the first for loop on line~\ref{alg:replace1:for1}, we add attack links from the argument that has just been created to all existing arguments for the IE. The next two for loops on respectively lines~\ref{alg:replace1:for2} and~\ref{alg:replace1:for3} ensure that all attack links that existing from and to the previous versions of the IE are also carried over to the new argument $A$. Finally, one line~\ref{alg:replace1:addarg}, the new argument is added to the set of arguments.

Let us explain this with an example. \todo{Marc}{Marc}{Add example}

>>>>

\subsection{Translation algorithms}

Now that we have formalized both a GRL model and a RationalGRL model, we develop algorithm in order to translate between these two models. Both of these two translation algorithms are straightforward, which is result of the fact that the two models are formalized in a very similar way.

\paragraph{GRL to RationalGRL Translation} We start with the translation algorithm from GRL to RationalGRL, which is shown in Algorithm~\ref{alg:translation:to-rationalgrl}. The translation algorithm takes a GRL model $GR=(IE, Act, R_{ActIE}, Link)$ (Def.~\ref{def:grl-model}) as input. On line~\ref{alg:translation:to-rationalgrl:args}, it collects all the elements of the GRL model into a set $Arg$, which represents the set of arguments of the RationalGRL model. On line~\ref{alg:translation:to-rationalgrl:att}, the set of attack links is initialized with an empty set, and the new RationalGRL model $(Arg, Att)$ is returned on line~\ref{alg:translation:to-rationalgrl:return}.

\begin{algorithm}[h]
  \caption{GRL to RationalGRL Translation}
  \label{alg:translation:to-rationalgrl}
  \begin{algorithmic}[1]
    \Procedure{$ToRationalGRL$}{$GRL$}
    \State $Arg \leftarrow IE\cup Act \cup R_{ActIE}\cup Link$\label{alg:translation:to-rationalgrl:args}
    \State $Att \leftarrow \emptyset$\label{alg:translation:to-rationalgrl:att}
    \State \Return $(Arg, Att)$\label{alg:translation:to-rationalgrl:return}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\paragraph{RationalGRL to GRL translation} The translation from a RationalGRL model to a GRL model is given in Algorithm~\ref{alg:translation:to-grl}. It is very similar to the previous algorithm, but then works in the other direction. In this case, arguments are unpacked and put in the corresponding GRL components. On line~\ref{alg:translation:to-grl:extension}, argumentation semantics is used to determine the \emph{preferred extension} (Def.~\ref{def:preferred-semantics}). This is a set containing all accepted arguments. The GRL model is then generated from all the accepted arguments. This is done by iterating over all arguments in the extension in line~\ref{alg:translation:to-grl:for}. The switch statement on line~\ref{alg:translation:to-grl:switch} then does a case distinction on the type of the arguments, and each case ensures the argument is put in the right GRL component. Finally, the algorithm returns a GRL model on line~\ref{alg:translation:to-grl:return}.

\begin{algorithm}[h]
  \caption{RationalGRL to GRL Translation}
  \label{alg:translation:to-grl}
  \begin{algorithmic}[1]
    \Procedure{$ToGRL$}{$RatGRL$}
    \State $E \leftarrow ComputeExtension(Arg,Att)$\label{alg:translation:to-grl:extension}
    \State $IE\leftarrow\emptyset$, $Act\leftarrow\emptyset$, $R_{ActIE}\leftarrow\emptyset$, $Link\leftarrow \emptyset$
    \For{$A\in E$}\label{alg:translation:to-grl:for}
      \Switch{$A.type$}\label{alg:translation:to-grl:switch}
          \Case{$IETypes$}
            \State $IE\leftarrow IE\cup \{A\}$
          \EndCase
          \Case{$Actor$}
            \State $Actor\leftarrow Actor\cup \{A\}$
          \EndCase
          \Case{$ActIE$}
            \State $ActIE\leftarrow ActIE\cup \{A\}$
          \EndCase
          \Case{$LinkTypes$}
            \State $Link\leftarrow Link \cup\{A\}$
          \EndCase
      \EndSwitch
    \EndFor
    \State \Return $(IE,Act,R_{ActIE}, Link)$\label{alg:translation:to-grl:return}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\paragraph{Valid RationalGRL model} While we have defined a notion of a \emph{valid GRL model} (Def.~\ref{def:valid-grl-model}), we have no  done so for a RationalGRL model yet. This is because it depends on the translation procedure. We define a RationalGRL model as valid if and only if the RationalGRL to GRL translation results in a valid GRL model. In this way, we do not have to reiterate all conditions on a GRL model, but use the translation algorithm.

\begin{definition}[Valid RationalGRL Model]
\label{def:valid-rationalgrl-model}
A RationalGRL model $RatGRL = (Arg, Att)$ (Def.~\ref{def:rationalgrl-model})
is a \emph{valid RationalGRL model} iff $ToGRL(RatGRL)$ (Algorithm~\ref{alg:translation:to-grl}) is a valid GRL model (Def.~\ref{def:valid-grl-model}).
\end{definition}

@@

In the previous subsection we formalized a \emph{static} representation of the RationalGRL framework. In this section we formalize the \emph{dynamics}. We do so by developing algorithms for applying argument schemes and critical questions in the context of a RationalGRL model (Def.~\ref{def:rationalgrl-model}). These algorithms are applied to RationalGRL models and produce new arguments and attack relations. We can then use argumentation semantics (Def.~\ref{def:preferred-semantics}) to compute sets of accepted arguments. The content of these arguments is then used to compute the resulting RationalGRL model, together with enabled and disabled GRL elements and their underlying arguments.

As discussed in Section~\ref{sect:overview}, the argument schemes and critical questions of Table~\ref{table:argument-schemes} fall into one of the following four categories:
\begin{itemize}
\item \emph{INTRO}: These algorithms add one or multiple new arguments to the RationalGRL model, without creating any attack links. The arguments are for new GRL elements or links.
\item \emph{DISABLE}: These algorithms add a single argument to the RationalGRL model, which attacks one of multiple existing arguments. The counterargument that is added does not correspond to any GRL element, but instead disables an existing one by attacking all arguments for it.
\item \emph{REPLACE}: These algorithms can be seen as a combination of \emph{INTRO} and \emph{DISABLE}. They add a new argument corresponding to a GRL element or link, and this new argument attacks a previous version of the element or link. 
\end{itemize}

@@

\paragraph{Rationale and example} Algorithm~\ref{alg:add-actor} takes one argument, namely the name of the actor $a$. On line~\ref{alg:as0:1} of the algorithm, a new (unique) id is minted as the identifier of the new actor, which is assigned with its corresponding name to the variable $actor$. On line~\ref{alg:as0:2} the actor is added as a new argument.

As an example, we can formalize the addition of actor \emph{Traffic Simulator} in Figure~\ref{fig:example-small2} as executing the algorithm $AS_0($Traffic simulator), which results in a RationalGRL model $RatGRL=(Arg,Att)$ with argument $(i, Actor, \text{Traffic simulator})\in Arg$.

\begin{algorithm}[h]
  \caption{AS1: Actor with id $i$ has resource $R$}\label{alg:add-resource}
  \begin{algorithmic}[1]
    \Procedure{$AS_1$}{$i, R$}
    \State $res\_id\gets mintId()$\label{alg:add-resource:id}
    \State $A_1\leftarrow (res\_id, Resource, R, AND)$\label{alg:add-resource:arg1}
    \State $A_2\leftarrow (mintId(),ActIE, i, res\_id)$\label{alg:add-resource:arg2}
    \State $Arg\gets Arg\cup \{A_1,A_2\}$\label{alg:add-resource:add-arg}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

@@

The algorithm itself runs as follows: on line~\ref{alg:add-resource:id}, a unique id is assigned to variable $res\_id$ (``resource identifier''). On line~\ref{alg:add-resource:arg1}, an argument for a resource with name $R$ and identifier $res\_id$ is assigned to $A_1$. On line~\ref{alg:add-resource:arg2}, a second argument $A_2$ is created, which is an argument for an Actor-IE relation (Def.~\ref{def:act-ie-relation}) between the input actor $i$ and the newly created element $res\_id$. Finally, on line~\ref{alg:add-resource:add-arg}, the arguments are added to the set of arguments $Arg$.

As an example, consider the addition of resource ``Car objects'' in Figure~\ref{fig:example-small2}, and suppose the actor is formalized as the argument $(i, Actor, \text{Traffic simulator})$. Adding the resource can then be formalized as executing $AS1(i, \text{Car objects})$, resulting in the addition of two arguments $(j, Resource, \text{Car object}, AND)$ and $(k, ActIE, i, j)$ which are added to the set of arguments $Arg$.

@@

\begin{algorithm}[h]
  \caption{AS5: Goal with id $i$ decomposes into task $T$}\label{alg:add-decomp}
  \begin{algorithmic}[1]
    \Procedure{$AS_5$}{$i, T$}
    \State $task\_id\gets mintId()$\label{alg:add-decomp:task-id}
    \State $A_1\leftarrow (task\_id, Task, T, AND)$\label{alg:add-decomp:arg1}
    \State $A_2\leftarrow (mintId(), Decomp, i, task\_id)$\label{alg:add-decomp:arg2}
    \State $Arg\gets Arg\cup \{A_1,A_2\}$\label{alg:add-decomp:add-arg}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

@@

As an example, suppose we are adding the fact that goal ``Generate traffic'' is decomposing into task ``Keep same cars'' (Figure~\ref{fig:example-small2}), and suppose the following argument for the goal exists already: $(i, Goal,\text{Generate traffic}, AND)$. Adding the decomposition can be formalized as executing $AS_5(i, \text{Keep same cars})$, and results in two new arguments: $(task\_id, Task, \text{Keep same cars}, AND)$ and $(k, Decomp, i, task\_id)$, which are both added to the set of arguments $Arg$.

@@

Consider for example the RationalGRL model in Figures~\ref{fig:examples:relevant-actor}, which consists of an actor and a generic counterargument. Let us reconstruct this model using the application of argument schemes and critical questions. Initially, the RationalGRL model $(Arg,Att)$ is empty, i.e. $Arg=Att=\emptyset$. Next, the algorithm $AS_0(\text{Development team})$ is executed, which results in $A_1=\{(0,Actor,\text{Development team}\})$. After this, critical question CQ0: ``Is the actor Development team relevant?'' is answered with ``No, because the professor develops the software''. We can formalize this as executing algorithm $CQ_0(0)$ (Algorithm~\ref{alg:actor-not-relevant}), which results in adding a generic counter argument $A_2=(1, GenArg, \text{CQ0: Professor develops software})$ and an attack link $(A_2,A_1)$. Note that the RationalGRL model now consists of two arguments, and that the preferred extension is $\{A_2\}$, which means that the resulting GRL model is empty, because generic arguments are not translated to GRL.

@@

Let us reconsider Figure~\ref{fig:examples:relevant-actor}, which we previously formalized as $RatGRL=(Arg,Att)$, where $Arg=\{A_1,A_2\}$, $A_1=\{(0,Actor,\text{Development team}\}), A_2=(1, GenArg, \text{CQ0: Professor develops software})$, and $Att=(A_2,A_1)$. Suppose we execute algorithm $Att(\{A_2\}, \text{Professor does not develop software})$. This results in $A_3=(2, GenArg, \text{CQ0: Professor does not develop software})$ and a new attack link $(A_3,A_2)$ (Figure~\ref{fig:examples:relevant-actor2}). Since the argument for the goal is now part of the preferred extension, it does show up in the resulting GRL model after the translation of Algorithm~\ref{alg:translation:to-grl}.

@@

\begin{algorithm}[h]
  \caption{CQ5c: Is the decomposition type of element $ie_i$ correct? No, it should be $X$ }\label{alg:replace1}
  \begin{algorithmic}[1]
    \Procedure{$CQ_{Decomp}$}{$ie_i, X$}
    \State $A \leftarrow ie_i$\label{alg:replace1:arg}
    \State $A.decomptype\leftarrow X$\label{alg:replace1:decompchange}
    \State $IEArgs\leftarrow IE_i\subseteq  Arg$\label{alg:replace1:ieargs}
    \For{$\{(A_i,A_j)\in Att\mid A_j\in IEArgs\}$}\label{alg:replace1:for1}
      \State $Att\leftarrow (A_i,A)$
    \EndFor
    \For{$\{(A_i,A_j)\in Att\mid A_i\in IEArgs\}$}\label{alg:replace1:for2}
      \State $Att\leftarrow (A,A_j)$
    \For{$A_i\in IEArgs$}\label{alg:replace1:for3}
      \State $Att \leftarrow Att \cup \{(A,A_i)\}$\label{alg:replace1:att}
    \EndFor
    \EndFor
    \State $Arg\leftarrow Arg \cup \{A\}$\label{alg:replace1:addarg}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\paragraph{Rationale and example} While the original critical question CQ5c is specific to the decomposition between a goal and a task, Algorithm $CQ_{Decomp}$~\ref{alg:replace1} is more generally applicable to the decomposition type of any IE, since all IEs have a decomposition type in their definition (Def.~\ref{def:ie}).

Let us go through this algorithm step by step. On line~\ref{alg:replace1:arg}, a new argument $A$ is created which is identical to original IE. Recall from Def.~\ref{def:notation} that we may refer to elements of an IE with the ``.'' notation. This is what we do on line~\ref{alg:replace1:decompchange}, where the decomposition type of the argument is changed to $X$. On line~\ref{alg:replace1:ieargs}, the set $IEArgs$ is assigned with all existing arguments for the input IE (recall the notation of $IE_i$ as the set of all IEs with id $i$, see Def.~\ref{def:notation}). The first two for loops on respectively lines~\ref{alg:replace1:for1} and~\ref{alg:replace1:for2} ensure that all attack links that existing from and to the previous versions of the IE are also carried over to the new argument $A$. Then, in the third for loop on line~\ref{alg:replace1:for3}, we add attack links from the argument that has just been created to all existing arguments for the IE. This ensures that all previous version of the IE are not part of the preferred extension, and as a result do not show up in the resulting GRL model. Finally, one line~\ref{alg:replace1:addarg}, the new argument is added to the set of arguments.

Let us explain this algorithm with the example in Figure~\ref{fig:examples:decomposition}. The initial RationalGRL model (before applying CQ10b), can be formalized as follows: $RatGRL=(Arg,Att)$, with $Arg=\{A_1,A_2,A_3,A_4,A_5\}$ and $Att=\emptyset$, where:
\begin{itemize}
\item $A_1=(0,Goal,\text{Simulate traffic},AND)$
\item $A_2=(1,Task,\text{Dynamic simulation},AND)$
\item $A_3=(2,Task,\text{Static simulation},AND)$
\item $A_4=(3,Decomp,0,1)$
\item $A_5=(4,Decomp,0,2)$
\end{itemize}

Suppose algorithm $CQ_{Decomp}(0, OR)$ is executing for this RationalGRL model. On lines~\ref{alg:replace1:arg} and \ref{alg:replace1:decompchange} a new argument $A_6=(0,Goal,\text{Simulate traffic},OR)$ is created. On line~\ref{alg:replace1:ieargs}, $IEArgs$ contains all arguments for element 0, which is $\{A_1\}$ (note this does not yet include argument $A_6$, since it is not yet added to the RationalGRL model). The first two for loops on lines~\ref{alg:replace1:for1} and~\ref{alg:replace1:for2} do not do anything in the case of our example, since $A_1$ is itself not attack or does not attack any other argument. The for loop on line~\ref{alg:replace1:for3} ensures all previous versions of the element with id 0 are now attacked, which means in our case the $(A_6,A_1)$ is added to $Att$. Finally, on line~\ref{alg:replace1:addarg} the new argument $A_6$ is added to $Arg$.
